---
title: "Math Performance in Relation to School and Social Factors"
author: "Isabel Gomez, Dhajanae Sylvertooth, Nicholas Hartman, Ritoban Kundu"
date: ""
bibliography: bib.bib
biblio-style: unsrt
output:
  pdf_document:
    pandoc_args: ["--natbib"]
---

\begin{center}
GITHUB: \url{https://github.com/igomez39/BIOS625}.
\end{center}

\section{Work Contributions}

Everyone: Report writing, Project planning \newline
Isabel: ggplot visuals \newline
Dhajanae:  ggplot visuals \newline
Nicholas: Data long format code, Gini Index bootstrap analysis, Cluster computing for GLS model \newline
Ritoban: Multiple Imputation of the Data, Longitudinal Data Analysis

\section{Introduction}

Academic proficiency during the formative years of schooling has widespread implications on future development [@first]. In this report, we focus on math performance in elementary school students over time, and study the relationships between math proficiency and several social and school characteristics. Using the Gini Index, nonparametric estimation, and longitudinal modeling techniques, we explore several different aspects of math proficiency, such as inequality and progress over time. 

The data for this analysis come from the Early Childhood Longitudinal Study (ECLS), which was a prospective cohort study of elementary school students from 2011-2016 [@ECLS]. Students were followed from kindergarten to fifth grade, and math performance was assessed each year. In addition, detailed surveys were conducted over time to collect data on a wide range of social factors. The public-use dataset contains information on over 26,000 variables for 970 schools and 18,174 individual students.

\section{Descriptive Plots}

We first made descriptive plots to visualize the demographics of the study sample and basic trends in the data. Most students in the cohort attended a public school, and both the teachers and parents in this sample tended to be highly educated. However, there was still enough diversity in these variables to study the differences across groups. The cohort included a well-balanced mix of students from urban, suburban, and rural areas, which is a strength of the study design. 

Based on plots of the average math scores over time, we observed little difference in math scores across public and private schools. The math scores also appeared to be very similar regardless of teacher education level. However, we did observe that students with highly-educated parents had the largest average math scores every year (see plots on the next page).  

```{r, echo=FALSE,warning=F,message=F}
setwd("//Users//nicholashartman//Downloads//Education Data")

library(tidyverse)
library(data.table)
library(gridExtra)
#Read Data
dat=as.data.frame(fread("Imputed_Data.csv",sep=","))
#####Visuals######
library(ggplot2)
library(treemapify)
attach(dat)
#par(mar = c(5.1,4.5,4.1,2.5))
#####School Type frequency table
g1=(ggplot(dat, aes(x = as.factor(Pub_Pri), fill = factor(Pub_Pri))) + geom_bar(width = 0.5) +
coord_flip() +
labs(
x = "School Type",
y = "Frequency",
fill = "School Type") +
scale_x_discrete(labels = c("Public", "Private")) +
theme(legend.position = "none") +
scale_fill_discrete(labels = c("Public", "Private")))
#####Teacher Education frequency table
g2=(ggplot(dat, aes(x = as.factor(Teacher_Education), fill = factor(Teacher_Education))) + geom_bar(width = 0.5) +
coord_flip() +
labs(
x = "Teacher Education (yrs)",
y = "Frequency",
fill = "Teacher Education") +
theme(legend.position = "none") +
scale_fill_discrete(labels = c("5 yrs", "6 yrs", "7 yrs")))
#####Parent1 & Parent2 Education frequency table
g3=(ggplot(dat, aes(x = factor(Parent1_Education), fill = factor(Parent1_Education))) + geom_bar(width = 0.5) +
coord_flip() +
labs(
x = "Parent Level of Education",
y = "Frequency",
fill = "Level of Education") +
scale_x_discrete(labels = c("8th grade or less", "9th-12th grade", "High School", "Voc/Tech Program", "Some College", "Bachelor's Degree", "Grad/ Professional", "Master's Degree")) +
theme(legend.position = "none",
plot.margin = margin(r = 20, l = 20, t = 5, b = 5),
legend.spacing = unit(1.0, 'cm'),
legend.key.size = unit(1, 'cm')) +
scale_fill_discrete(labels = c("8th grade \n or less", "9th-12th grade", "High School", "Voc/Tech Program", "Some College", "Bachelor's Degree", "Grad/ Professional", "Master's Degree")))
#####Region Frequency Table
dat$Locale=factor(dat$Locale)
g4=(ggplot(dat, aes(x = factor(Locale), fill = factor(Locale))) +
geom_bar(width = 0.5) +
coord_flip() +
labs(
x = "Locale",
y = "Frequency",
fill = "Locale") +
theme(legend.position = "none") +
scale_x_discrete(labels = c("City", "Suburb", "Town", "Rural")) +
scale_fill_discrete(labels = c("City", "Suburb", "Town", "Rural")))
grid.arrange(g1,g2,g3,g4)
```

```{r,echo=FALSE,warning=FALSE,message=F,fig.height=4}

#Read Data
setwd("//Users//nicholashartman//Downloads//Education Data")
edu_data=as.data.frame(fread("Data_Long_Format.csv"))

#COLOR BLIND FRIENDLY PALETTE
cbp1 <- c("#999999", "#E69F00", "#56B4E9", "#009E73",
          "#000000", "#0072B2", "#D55E00", "#CC79A7")

###PREPPING SCHOOL DATA
edu_data <- edu_data %>% group_by(Teacher_Education) %>% mutate(avg_math = mean(Math_Score)) %>% filter(is.na(Teacher_Education)==FALSE, is.na(Math_Score) ==FALSE, Math_Score != -9, Teacher_Education != -9, Parent1_Education != -9, Parent2_Education != -9)

#Grouping by school

edu_data_school <- edu_data %>% filter(is.na(School_ID)==FALSE, is.na(Math_Score) ==FALSE, Pub_Pri != -1) %>%
  group_by(School_ID)  %>% mutate(avg_math_school = mean(Math_Score))

edu_data_school$pub_pri = as.factor(edu_data_school$Pub_Pri)
levels(edu_data_school$pub_pri) <- c("Public","Private")

test = data.frame(tchr_edu = unique(edu_data$Teacher_Education), avg_math = unique(edu_data$avg_math))

edu_data_school$Teacher_Education = as.factor(edu_data_school$Teacher_Education)
levels(edu_data_school$Teacher_Education) <- c("Bachelors Degree or Less", 
                                        "Master's Degree", "Advanced Degree")

#AVERAGE MATH SCHOOL BY YEAR 
edu_data_school <- edu_data_school %>% group_by(Dates) %>% 
  mutate(avg_math_yr = mean(Math_Score))

#RECODING SCHOOL YEAR TO REPRESENT GRADE LEVEL
edu_data_school$Dates = as.factor(edu_data_school$Dates)
levels(edu_data_school$Dates) <- c("K","1","2","3","4","5")

#AVERAGE MATH SCORE BY TEACHER EDUCUATION
edu_data_school <- edu_data_school %>% group_by(Teacher_Education, Dates) %>% 
  mutate(avg_math_yr_tchr = mean(Math_Score))

edu_data_school$Parent1_Education = as.factor(edu_data_school$Parent1_Education)
levels(edu_data_school$Parent1_Education) <- c("8th grade or less",
                                               "9-12th grade", 
                                               "HS Diploma/Equivalent",
                                               "Vocational/Technical School", 
                                               "Some College",
                                               "Bachelors Degree", "Graduate/Professional School", "Masters", "PhD or Professional Degree")
edu_data_school <- edu_data_school %>% group_by(Parent1_Education, Dates)%>%
  mutate(avg_math_yr_parent = mean(Math_Score))


#DIFFERENCE IN AVERAGE MATH SCHOOL
edu_data_school <- edu_data_school %>%
  group_by(Dates) %>%
  mutate(diff_math_avg = avg_math_yr - lag(avg_math_yr, default = 0))

g1=(ggplot(edu_data_school, aes(x = Dates, y = avg_math_yr)) +
  geom_point() +
  facet_wrap(as.factor(edu_data_school$pub_pri))+
  labs(x = "Grade Level" , y = "Average Math Score") +
  theme_classic())

g2=(ggplot(edu_data_school, aes(x = Dates, y = avg_math_yr_tchr, col =Teacher_Education)) +
  geom_point() +
  labs(x = "Grade Level" , y = "Average Math Score",colour="Teacher Education") +
  theme_classic())

g3=(ggplot(edu_data_school, aes(x = Dates, y = avg_math_yr_parent, col = Parent1_Education)) +
  geom_point() +
  labs(x = "Grade Level" , y = "Average Math Score",colour="Parent Education") +
  theme_classic() +
  scale_colour_manual(values = cbp1))

grid.arrange(g2,g3,nrow=2)
```

```{r,echo=FALSE,fig.height=2}
g1
```

\section{Math Score Inequality}

\noindent We studied the amount of inequality in math performance, and how this changes over time within public and private elementary schools. To accomplish this, we used the Gini Index, which is a well-known measure of inequality and is commonly used to describe wealth inequality. The formula for the Gini Index is [@Inequality]:

$$\textrm{Gini}=\frac{\sum_{i=1}^n \sum_{j=1}^n |x_i-x_j|}{2n^2\bar{x}},$$
\noindent where $x$ represents the math score. Higher Gini Index values correspond to greater inequality in the math scores. For this analysis, we used a re-scaled version of the math scores since $x$ must be positive in the Gini Index formula. There are two main computational challenges that arise in Gini Index estimation [@Inequality]:

\begin{enumerate}
\item Due to a lack of theoretical formulas for the Gini Index standard error, nonparametric bootstrapping is needed to estimate the uncertainty in the Gini Index. 
\item The Gini index involves many pairwise comparisons for large datasets.
\end{enumerate}

We addressed the challenges of (2) by using a vectorized implementation of the Gini Index from the reldist package in R [@reldist]. Then, to further speed up the resampling in (1), we used parallel computing. For each type of school (Public vs. Private) and for each timepoint, we resampled the corresponding data with replacement 10,000 times and computed the Gini Index on each of these bootstrap samples. A 95\% confidence interval for each point estimate was computed from the quantiles of the bootstrap distribution. The figure below shows the Gini Index estimates and bootstrap 95\% confidence intervals over time, stratified by school type:

```{r, echo=FALSE,fig.width=12}
setwd("//Users//nicholashartman//Downloads//Education Data")

#Read from cluster-generated files
ests=as.matrix(read.csv("ests.csv")[,-1])
boots=as.matrix(read.csv("boots.csv")[,-1])
lower=as.matrix(read.csv("lower.csv")[,-1])
upper=as.matrix(read.csv("upper.csv")[,-1])
ggdata=data.frame(ests=c(ests[1,],ests[2,]),lower=c(lower[1,],lower[2,]),time=c(1:6,1:6),
                  upper=c(upper[1,],upper[2,]),school=c(rep("Public",6),rep("Private",6)))
g=ggplot(ggdata,aes(x=time,y=ests,group=school,color=school))+geom_line()+geom_point()
g=g+geom_errorbar(aes(ymin=lower, ymax=upper))+theme_classic()
g=g+ylab("Gini Index")+xlab("Cohort Assessment")
g=g+scale_x_continuous(breaks=1:6,labels=c("Spring 2011","Spring 2012","Spring 2013",
                                            "Spring 2014","Spring 2015","Spring 2016"))
g=g+theme(text=element_text(size=15))+labs(color="School Type")
g
```

\noindent We find that the Gini Index decreases over time within both public and private schools. One interpretation of this result is that the kindergarten students with lower scores were able to catch up to the high-performers by fifth grade. We also observed that there is more inequality in math scores within public schools compared to within private schools. The magnitude of this difference remained fairly constant over time, and since the 95\% confidence intervals do not overlap, we can conclude that these differences are statistically significant. 

\section{Longitudinal Models}

To study the relationships between social factors and math performance over time, we used a Generalized Least Squares (GLS) approach. We chose to use GLS over mixed models because we are mainly interested in the marginal associations. No structure was assumed for the covariance matrix of scores over time because it resulted in a substantially better model fit (and AIC) compared to other covariance structures (such as Exponential or Hybrid Structure). We specified the mean model as:

\begin{align*}E[\textrm{Score}_{ij}]&=\beta_0+\beta_1\textrm{Time}_{ij}+\beta_2 I(\textrm{School}_i=\textrm{Private})+\beta_3 \textrm{Teacher\_Education}_i+\beta_4 \textrm{Parent\_Education}_i\\
 &+\beta_5 I(\textrm{Locale}_i=\textrm{Rural})+\beta_6 I(\textrm{Locale}_i=\textrm{Town})+\beta_7 I(\textrm{Locale}_i=\textrm{Suburban}) \\
 &+ \beta_8  I(\textrm{School}_i=\textrm{Private}) \times \textrm{Time}_{ij}+\beta_9 \textrm{Teacher\_Education}_i \times \textrm{Time}_{ij} \\
 &+ \beta_{10} \textrm{Parent\_Education}_i \times \textrm{Time}_{ij} +\beta_{11} I(\textrm{Locale}_i=\textrm{Rural}) \times \textrm{Time}_{ij} \\
 &+\beta_{12} I(\textrm{Locale}_i=\textrm{Town}) \times \textrm{Time}_{ij} + \beta_{13} I(\textrm{Locale}_i=\textrm{Suburban}) \times \textrm{Time}_{ij}.
 \end{align*}

Given that a nontrivial amount of data was missing, we performed multiple imputation with chained equations using the mice package in R [@mi]. The data were imputed using the predictions from predictive mean matching, logistic, proportional odds, and multinomial models for continuous, binary, ordinal, and categorical data respectively. One hundred imputed datasets were created and the GLS model was fit to each one simultaneously using cluster computing. This reduced the computational time from approximately 8 hours to 5 minutes. The estimates were then combined according to Rubin's multiple imputation rules [@LR]. Any model that did not converge was discarded from the analysis. A forest plot of the point estimates and 95\% confidence intervals is shown:

```{r, echo=FALSE, fig.height=5, fig.width=10}
#Run on local computer to make plot
setwd("//Users//nicholashartman//Downloads")
GLS=read.csv("GLS.csv")[,-1]
GLS_SE=read.csv("GLS_SE.csv")[,-1]
low=GLS-1.96*GLS_SE
up=GLS+1.96*GLS_SE
label=c("Intercept","Teacher Education","Time","Parent Education","Suburb","Town","Rural",
            "Private School","Teacher Education x Time","Parent Education x Time","Suburb x Time",
            "Town x Time","Rural x Time","Private School x Time")
GLS_data=data.frame(GLS,low,up,label)
GLS_data=GLS_data[14:2,]
GLS_data=GLS_data[c(1:11,13,12),]
GLS_data$label_num=c(1:13)
g=ggplot(GLS_data)+geom_point(aes(x=label_num,y=GLS))+coord_flip()
g=g+geom_hline(yintercept=0,col="green")+theme_classic()
g=g+geom_errorbar(aes(x=label_num,ymin=low,ymax=up))
g=g+scale_x_continuous(breaks=GLS_data$label_num,label=GLS_data$label)
g=g+ylab("Estimate")+xlab("")
g
```

Based on the forest plot above, we see a strong positive time trend in the cohort's math scores, as expected. Children who have parents with a higher education perform significantly better in kindergarten, but have significantly less improvement in their scores over time. Schools that are located in rural areas, towns, or suburban areas all have significantly higher math scores at baseline compared to urban schools, but we see no significant differences in math score improvement over time. 

\section{Discussion}

The results of this analysis have important policy implications. For example, future work is needed to uncover why public schools have more inequality in math scores, and what interventions can be used to reduce these education gaps. In our longitudinal modeling, we observed that having a highly-educated parent is significantly associated with higher scores in kindergarten. Schools and policymakers may attempt to identify specific factors that are driving this association and provide all students with the same resources. 





